---
title: "README"
crossref: 
  fig-title: Figuure
  tbl-title: Table
format: 
  gfm: 
    html-math-method: mathjax
    comment: "#>"
    collapse: true
    toc: true
tbl-cap-location: top
fig-cap-location: bottom
prefer-html: true
keep-yaml: true
---

<!-- README.md is generated from README.qmd. Please edit README.qmd, then render README.md using `quarto render README.qmd`. -->
<!-- use `quarto render README.qmd` to build README.md prior to committing to keep README.md up-to-date-->
<!-- don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tmbGMACS

<!-- badges: start -->
<!-- badges: end -->

tmbGMACS is an [R](https://www.r-project.org/) package that, when completed, will provide a TMB-based implementation of the 
Generalized Model for Assessing Crustacean Stocks ([GMACS](https://github.com/GMACS-project/GMACS_Assessment_code)) developed 
using AD MODEL BUILDER ([ADMB](http://www.admb-project.org)). TMB ([Template Model Builder](https://kaskr.github.io/adcomp)) is 
an R package similar to ADMB that provides the ability to develop and run C++-based statistical models using 
automatic differentiation in the R framework.

## Installation

You can install the development version of tmbGMACS from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wStockhausen/tmbGMACS")
```

```{r links, include=FALSE}
  bbrkc = "https://meetings.npfmc.org/CommentReview/DownloadFile?p=b98b90b2-88ab-43c2-9487-c12cdb4e0a25.pdf&fileName=BBRKC%20SAFE%202022%20Final.pdf";
  snow_crab = "https://meetings.npfmc.org/CommentReview/DownloadFile?p=fca55335-ad34-4896-9b1e-4c09aa8342ce.pdf&fileName=EBS%20Snow%20SAFE%20FINAL.pdf";
```

## Introduction 

The purpose of tmbGMACS is to implement a population model for a fished 
crustacean stock and fit it to fishery and survey data to estimate various 
population processes (e.g., recruitment, growth, and mortality) and population size 
in order to determine sustainable fishing limits and targets. The ADMB version of GMACS 
is currently used to provide fishery stock assessment information to the 
North Pacific Fishery Management Council ([NPFMC](https://www.npfmc.org)) for a number 
of crab stocks in the Bering Sea and Aleutian Islands (see, for example, the stock assessment 
reports for [Bristol Bay red king crab](`r bbrkc`) and [snow crab](`r snow_crab`)).

Here, the TMB model is referred to as "gmacs" while the R package is referred to "tmbGMACS". The 
R package provides functions to set up data inputs to gmacs and associated likelihood functions, 
select functional forms for various population processes, 
define estimable parameters and associated priors or penalties, run the model, 
create plots and tables from a fitted model, and compare multiple models.

## Population model description 

The gmacs population model integrates the abundance of individuals in various discrete 
population categories over time based on 
assumptions regarding population processes consisting of recruitment, growth, maturation, 
natural mortality, and fishing mortality. Typical population categories might be 
sex and size, but it is possible to define an arbitrary number of categories (e.g., 
region, sex, maturity state, shell condition, and size). Although crustaceans cannot typically 
be aged, age could be included as a population category if population processes such as 
maturation were thought to have a strong age dependence.

Population abundance in gmacs is represented as vector, $\mathbf{n}$, with elements $n_i$, where $i$ is 
a 1-dimensional index $i$ combining time (year and season) and levels across the population categories. The temporal nature 
of the calculations is more easily described, though, by differentiating between time and the population categories, 
so $\mathbf{n_t}$ is a vector representing the abundance in each category at time $t$ (i.e., a vector with elements 
$\mathbf{n_t}_c$):


$$\frac{d\mathbf{n_t}}{dt} = \left[ \mathbf{T_t}-(\mathbf{M_t}+\mathbf{F_t}) \right]*\mathbf{n_t}$$ {#eq-1}


where $\mathbf{T_t}$ is a $c$ x $c$ matrix representing transition rates between population categories (e.g., growth or movement) at time $t$, $\mathbf{M_t}$ is a similarly-sized diagonal matrix representing natural mortality at time t, and $\mathbf{M_t}$ is a similarly-sized diagonal matrix representing fishing mortality at time $t$. For gmacs, rates within a (seasonal) timestep are assumed to be constant. Thus, @eqn-1 could be integrated to obtain

$$\mathbf{n_{t+1}} = e^{\mathbf{A_t} \delta t}*\mathbf{n_t}$$ {#eq-2}
 
where $\delta t$ is the time step from $t$ to $t+1$, and $\mathbf{A_t}=\left[ \mathbf{T_t}-(\mathbf{M_t}+\mathbf{F_t}) \right]$. In practice, when $\mathbf{T_t}$ represents growth processes, these are regarded as discrete-time processes with a season length of 0, so that 

$$ \mathbf{n_{t+1}} = \mathbf{T_t}*\mathbf{n_t} $$ {#eq-3}
 
for seasons in which growth occurs and 

$$ \mathbf{n_{t+1}} = e^{-\left[ \mathbf{M_t}+\mathbf{F_t} \right]*\delta t}*\mathbf{n_t}$$ {#eq-4}

for seasons with non-0 lengths in which mortality occurred.

Recruitment, the addition of new individuals to the model population categories, is also considered a discrete-time process associated with a season length of 0, so 

$$ \mathbf{n_{t+1}} = \mathbf{R_t}+\mathbf{n_t}$$ {#eq-5}

when recruitment occurs, where $\mathbf{R_t}$ is a vector of length $c$.

Let $<\mathbf{R_s}>$ represent the equilibrium (i.e., longterm mean) recruitment in (within-year) season $s$, and similarly for $<\mathbf{T_s}>$ and $\mathbf{Z_s}=\mathbf{M_s}+\mathbf{F_s}$. The equilibrium population size $<\mathbf{n_s}>$ in season $s$ is then determined by the solution to 

$$<\mathbf{n_s}> = $$ {#eq-6}

## Model dimensions 

Model dimensions consist of year (**y**), season (**s**), and a set of categories ({**c**}) into which the population is divided (e.g., sex and size). Levels within a given population category can be nested within another category (e.g., different size classes might be defined for males and females if they exhibited substantially different growth patterns). For a given year, population 
abundance is integrated across seasons within the year and different processes can occur in different seasons. 

Dimensions are defined using the function \code{[createSparseDimsMap]("man/createSparseDimsMap.Rd")}. As an example (@tbl-Ex1), the following creates a set of dimensions encompassing year (**y**), season(**s**), sex (**x**), and size (**z**), with different size classes for males and females. Note that the "dimension name(s)" are added to the vector or list defining the levels as an attribute "dmnms".

```{r ex1}
library(tmbGMACS);
#> Create the individual dimensions as vectors or lists (for nested dimensions)
#> --assign the attribute "dmnms", the dimension name(s) (plural if nested)
#> ----model years
vYs = 2001:2005;                  attr(vYs,"dmnms")<-"y";
#> ----model seasons
vSs = c("fall","spring");         attr(vSs,"dmnms")<-"s";
#> ----model sexes
vXs = c("MALE","FEMALE");         attr(vXs,"dmnms")<-"x";
#> ----model sizes, nested with sexes
vZs = list(  MALE=seq(40,60,5),
           FEMALE=seq(25,50,5));  attr(vZs,"dmnms")<-c("x","z");
#>--create the dimensions "map"
dms = tmbGMACS::createSparseDimsMap(y=vYs,s=vSs,x=vXs,z=vZs);
```

```{r}
#| label: tbl-Ex1
#| tbl-cap: "Example dimensions map."
#| eval: !expr '(require(kableExtra))'
#| echo: false
dms |> kableExtra::kbl(booktabs=TRUE);
```

A dimensions map has non-standard attributes: "dmnms", "dmlvs", and "dmlns". 
  * "dmnms" is a character vector with the dimension names.
  * "dmlvs" is a character vector with the unique levels for each dimension.
  * "dmlns" is a named integer vector with the length (number of unique levels) of each dimension
  
The first column in the model dimensions map is simply the row number associated with the combination of 
unique dimension levels the row reflects. It also provides the index into the population abundance 
vector for that combination of dimensions (and thus allows the extraction of the associated dimension levels given 
the index value).

Most model quantities are represented as vectors, and thus have associated dimensions maps in order to be able to identify the dimension levels associated with each index value.

## Aggregating over model dimensions

Data fit in gmacs typically represent quantities aggregated over some combination of population categories, so it is necessary to determine the mappping from model indices to data indices. The function [createAggregatorMap](man/createAggregatorMap.Rd) provides the required mapping of 1-d indices from one dimensions map to another, as illustrated in the following example (@tbl-Ex2) using the dimensions map \code{dms} from the previous example:

```{r ex2}
#> Example of how to create an aggregator map using "dms" from above
#> to aggregate by year (y) and sex (x) over season (s) and size (z). 
#>    Create dimensions map with only y and x
  aggDims = tmbGMACS::keepDims(dms,keep=c("y","x"));#--keeps only y and x dimensions, recreates 1-d vector index
#>    Create aggregator map from y,x,s,z to y,x
  aggMap  = tmbGMACS::createAggregatorMap(dms,aggDims,keepOrigDims=TRUE);
```

*j* is the vector index for aggDims, *i* is the vector index for dms.
```{r}
#| label: tbl-Ex2
#| tbl-cap: Example aggregator map (from indices i to indices j).
#| eval: !expr 'require(kableExtra)'
#| echo: false
aggMap |> kableExtra::kbl(booktabs=TRUE);
```
